---
title: "Preparing Data for Interpolation"
author: "Christopher Prener, Ph.D."
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing Data for Interpolation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(areal)
library(dplyr)
library(sf)

data(ar_stl_asthma, package = "areal")

data(ar_stl_race, package = "areal")

data(ar_stl_wards, package = "areal")
```

Depending on the state of your spatial data, they may require some cleaning and modification before interpolation. The types of issues to address pre-interpolation fall into a few distinct categories:

1. Data are not in the right format
2. Data are not in the right coordinate system
3. Data have variable name conflicts
4. Data have too many variables or observations and therefore must be subset

Each of these conditions will be discussed below. The following examples assume:

```{r}
library(areal)
library(dplyr)   # data wrangling
library(sf)      # spatial data operations

# load data into enviornment
race <- ar_stl_race
asthma <- ar_stl_asthma
wards <- ar_stl_wards

# create example data - non-spatial data
asthmaTbl <- ar_stl_asthma
st_geometry(asthmaTbl) <- NULL

# create example data - wrong crs
race83 <- st_transform(race, crs = 4269)
```

## Validating Data
`areal` has a built-in tool for data validation, `ar_validate()`, that provides an excellent starting place for ensuring your data are ready to interpolate. The validation process covers the first three issues listed in the introduction. It can be run in a simple format:

```{r validate-simple}
ar_validate(source = asthma, target = wards, varList = "ASTHMA", method = "aw")
```

If `ar_validate()` returns a `FALSE` value, it can also be run in a verbose manner that returns a detailed tibble:

```{r validate-verbose}
ar_validate(source = asthma, target = wards, varList = "ASTHMA", method = "aw", verbose = TRUE)
```

Use the `verbose = TRUE` output as a checklist to address issues before interpolating your data.

## Format Issues
Data need to be loaded as `sf` objects in `R`. If `ar_validate()` returns `FALSE` for the first condition, data are not stored as `sf` objects:

```{r validate-non-sf}
ar_validate(source = asthmaTbl, target = wards, varList = "ASTHMA", method = "aw", verbose = TRUE)
```

### External Spatial Data
If you have external data, using `st_read()` is the best way to load them:

```r
df <- sf::st_read("data.shp", stringsAsFactors = FALSE)
```

The `st_read()` function supports a variety of the most common data sources.

### sp Data
If you are working with data that is an `sp` object in `R`, the `sf` package has a function `st_as_sf()` can be used to convert the object to `sf`:

```r
sf_object <- sf::st_as_sf(sp_object)
```

### Data From tidycensus and tigris
If you are using either the [`tidycensus`](https://walkerke.github.io/tidycensus/) or [`tigris`](https://github.com/walkerke/tigris) packages to access spatial data, there are options for downloading data as `sf` objects as well. In `tidycensus`, the `geometry = TRUE` option should be used. In `tigris`, the `class = "sf"` option should be used.

```r
stl_race <- tidycensus::get_acs(geography = "tract", state = 29, county = 510, year = 2017, 
                                table = "B02001", output = "wide", geometry = TRUE)
stl_tracts <- tigris::tracts(state = 29, county = 510, class = "sf")
```

### Tabular Data
If you have tabular data, you will need to merge them with a shapefile or other spatial data set that contains the geometry for the corresponding spatial features. For American users, a potentially common scenario here will be table of census geography (such as tracts) that lacks the geometry for those features. These can be downloaded using [`tigris`](https://github.com/walkerke/tigris) and then combined using [`dplyr`](https://dplyr.tidyverse.org). The following example assumes the tract data contains an identification number column named `GEOID` with the appropriate tract identifiers for St. Louis:

```r
stl_tracts <- tigris::tracts(state = 29, county = 510, class = "sf")

tract_data <- dplyr::left_join(stl_tracts, asthmaTBL, by = "GEOID")
```

## Coordinate Systems
Two coordinate system rules are enforced by `ar_validate`: both the source and the target data must be in the same [coordinate system](https://geocompr.robinlovelace.net/spatial-class.html#crs-intro) and that coordinate system must be a projected coordinate system. The `sf` package contains a function named `st_crs()` for previewing the current coordinate system of your data:

```{r race-crs}
st_crs(race83)
```

The `EPSG` value `4269` refers to a specific type of coordinate system known as as geographic coordinate system. These data use latitude-longitude values, which vary in length and thus cannot be used to calculate area (a key component of carrying out areal interpolations). These data would fail the `ar_validate()` process:

```{r validate-non-matching-crs}
ar_validate(source = race83, target = wards, varList = "TOTAL_E", method = "aw", verbose = TRUE)
```

These data fail both the matching CRS validation and the planar CRS validation because the `race` and `wards` data are in two different coordinate systems. We've already seen the CRS data for `race` above so we'll look at the `CRS data for `wards`:

```{r wards-crs}
st_crs(wards)
```

If these data fail that process, `st_crs()` can be used to identify whether the source or the target (or both) are in an inappropriate coordinate system. The `EPSG` value `26915` refers to another type of coordinate system known as as projected coordinate system (or "planar" data). Further, `26915` is a particular type of projected coordinate system known as the [Universal Transverse Mercator](https://en.wikipedia.org/wiki/Universal_Transverse_Mercator_coordinate_system) coordinate system. This is a good option for using `areal` because it is available around the world. There are a variety of tools for identifying the appropriate UTM zone, including this excellent [interactive map](https://mangomap.com/robertyoung/maps/69585/what-utm-zone-am-i-in-#). Once you have an EPSG zone, the website [epsg.io](https://epsg.io) can be used to search for the appropriate `EPSG` value.

In this case, the `wards` data are in an appropriate system but the `race` data need to be transformed. The `st_transform()` function can be used to modify them to an appropriate coordinate system:

```{r transform-crs}
raceFixed <- st_transform(race83, crs = 26915)
```

Once this is done, these data should pass the `ar_validate()` process:

```{r validate-matching-crs}
ar_validate(source = raceFixed, target = wards, varList = "TOTAL_E", method = "aw", verbose = TRUE)
```

## Variable Conflicts
If a variable does not exist in the source data, the validation process will fail:

```{r}
ar_validate(source = race, target = wards, varList = "TOTAL", method = "aw", verbose = TRUE)
```

In this case, the fix is as easy as correcting the typo in our `ar_validate()` call - the variable `TOTAL` should really be `TOTAL_E`:

```{r}
names(race)
```

Another common problem is that a variable exists in both our source and target data. For instance, we could add a `TOTAL_E` variable to our target data:

```{r}
wardsVar <- mutate(wards, TOTAL_E = seq(1:28))

ar_validate(source = race, target = wardsVar, varList = "TOTAL_E", method = "aw", verbose = TRUE)
```

Now, we fail the validation process because there is a conflict between the source and target data - `TOTAL_E` exists in both data sets. We can use the `select()` function from `dplyr` to remove the offending variable from the target data before proceeding:

```{r}
wardsFixed <- select(wardsVar, -TOTAL_E)

ar_validate(source = race, target = wardsFixed, varList = "TOTAL_E", method = "aw", verbose = TRUE)
```

Another option would be to rename the column using the `rename()` function from `dplyr` or by using another approach for renaming columns. Either option will work for `areal`'s purposes.

## Subsetting
One cautionary note when using `areal` is that the validation process will not check to see if too many variables or observations exist in the data. Cleaning the data ahead of time so that only the relevant observations exist can help improve performance - if you have tract data for an entire state but only need a particular city, everything will go faster if you subset these data using the `filter()` function from `dplyr`:

```r
countyData <- filter(stateData, COUNTY == 510)
```

Having too many columns poses an organizational problem more than anything else. We strongly encourage users to limit their interpolated data to only the needed values. For example, the `wards` data contains an `AREA` column whose units we don't know and an `OBJECTID` column that is an artifact from its creation using an ESRI product:

```{r}
names(wards)
```

We can use the `select()` function from `dplyr` to remove these before interpolation since we don't need them:

```{r}
wardsSubset <- select(wards, -OBJECTID, -AREA)
```
